---
title: "Cost-effectiveness analysis for influenza vaccination with uncertainty"
toc: no
---

<div style="text-align: right">
[Home page](index.html)
</div>

```{r libs, cache=FALSE, echo=FALSE, results = "hide", warning = FALSE}
set.seed(2020)

knitr::opts_chunk$set(fig.cap='', fig.align="center", message=FALSE, warning=FALSE, fig.height = 6, fig.width = 10)
```

This model was developed by Philippe Beutels, Joke Bilcke and Lander Willem for illustration and teaching purposes only. Please make contact if you would like to use this model in any public way.

In the field of economic evaluation in general, it could be applicable to any intervention with effectiveness realised within one year (like many curative interventions). In the field of infectious disease, it would only be applicable to influenza, and only if vaccination targets a small proportion of the population that does not at the same time form a core transmitter group of the virus (e.g. it is unlikely to be suitable to model childhood influenza vaccination). Note that for most applications, practical use of this model would likely entail having separate, though similarly simple decision trees to obtain the unit cost estimates (implying that these cells would be intermediary outcomes, and that there would be additional input cells for each relevant stage of disease (for instance for the number of consultations, the unit cost per consultation, medication use, hospital days etc. for cases of pneumonia, and all other relevant disease stages)).

Note that the final output is independent of coverage and target group size, if fixed administration costs are set to 0. This is due to the exclusion of herd immunity effects in this model.  

For further information, questions or comments please contact us.

# Set working directory
The first step when creating a new R file is to specify a working directory. You need to tell the R session where you want it to work. To do this you use the command to "set the working directory" or `setwd()`. 

You can label this working directory as 'home', and have other locations stored for plotting or for getting data. This location should be an existing folder on your computer. For example

```{r results = "hide", eval = FALSE}
home <- "~/Documents/Modelling_intro/"     # on OS X
home <- "C:\\Documents\\Modelling_intro\\" # on windows
setwd(home)
```

# Package(s)

If BCEA package is not installed yet, you can do so by
```{r results = "hide", eval=FALSE}
if(!"BCEA" %in% installed.packages()){
        install.packages('BCEA')
}
```

Once installed, you have to load the package(s) at the start of the R file. This is done using 

```{r results = "hide"}
library(BCEA)
```

# Specific uncertainty distributions
For some parameters, the sample size is rather small or they should be strictly positive, hence other distribution might be preferred. Let's first define some help functions to facilitate sampling using a given mean, se, proportion, samples size or confidence interval:

```{r}
sample_gamma <- function(nsamples, mean, se){
  rgamma(nsamples, shape = mean^2 / se^2, rate = mean / se^2)
}

sample_beta_event <- function(nsamples, events, sample_size){
  return(rbeta(nsamples, shape1 = events, shape2 = sample_size - events))
}

sample_beta_mean <- function(nsamples, mean, se){
  alpha = mean^2 * (1 - mean) / se^2 
  beta  = alpha * (1 - mean) / mean
  return(rbeta(nsamples, shape1 = alpha, shape2 = beta))
}
```

# Goal and input

The goal is to calculate the cost-effectiveness of a vaccination program for the elderly against flu, taking into account parameter uncertainty as specified below:

```{r results = "hide"}

# set number of samples to include in the uncertainty analysis
nsamples <- 5000

# set the random-number-generator seed to get exactly the same results each time you run the code
set.seed(5698) 

# population details
targetgroup     <- 2500000 								# people>65 years
life_expectancy <- sample_gamma(nsamples,12,2)		# in years
discount_rate   <- 0.015

set.seed(5698) 
xx <- rbeta(10,4,96)
set.seed(5698) 
yy <- sample_beta_event(10,4,100) 	

?rbeta
# disease burden
force_of_infection      <- rbeta(nsamples,4,96) #sample_beta_event(nsamples,4,100) 	
p_hosp <- rbeta(nsamples,10,90) #sample_beta_event(nsamples,10,100)
case_fatality_ratio      <- rbeta(nsamples,30,970) # sample_beta_event(nsamples,30,1000)

# disease-related costs
cost_per_hosp_case     <- rnorm(nsamples,2000,100) # sample_gamma(nsamples, 2000, 100)  # euro
cost_per_non_hosp_case <- rnorm(nsamples,100,5) #sample_gamma(nsamples, 100, 5) 	  # euro

# program details
uptake_program                <- 0.80
vaccine_efficacy              <- rnorm(nsamples,mean=0.6,sd=(0.7-0.5)/(2*1.96)) # sample_beta_mean(nsamples, mean = 0.6, se = (0.7-0.5)/(2*1.96))
vaccine_price_per_dose        <- 10 			                       # euro
admin_cost_per_dose           <- rnorm(nsamples,50,4) #sample_gamma(nsamples, 50, 4)   # euro
fixed_program_cost           <- 0 			                         # euro

# QALY
QALYloss_hosp_case    <- sample_beta_mean(nsamples, 0.018, 0.0018)
QALYloss_non_hosp_case <- sample_beta_mean(nsamples, 0.0082, 0.0018)

```

# Health Technology Assessment: step-by-step

Now, propagate the uncertainty from the input parameters of the flu model into the outcome of your flu model. To start the health technology assessment, we calculate the vaccine cost per dose and the treatment cost per case, taking into account the proportion of cases that lead to hospitalization, the QALY lost per case, and the discounted life expectancy:
```{r }
vaccine_cost_per_dose     <- vaccine_price_per_dose + admin_cost_per_dose
treatment_cost_per_case   <- cost_per_hosp_case * p_hosp + 
                              (1-p_hosp) * cost_per_non_hosp_case
QALYloss_per_case        <- QALYloss_hosp_case * p_hosp +        
                               QALYloss_non_hosp_case * (1-p_hosp)

life_expectancy_disc <- rep(NA,nsamples)
for (i in 1:nsamples){
        life_expectancy_disc[i] <- sum(1/((1+discount_rate)^(0:(life_expectancy[i]-1))))
}

```


Calculate the burden of disease with the program in place, along with the associated medical and program costs:
```{r eval = FALSE}
### Similar to the previous tutorial: "HTA Flu (basic)"
```

```{r include = FALSE}
# Program
population_at_risk         <- targetgroup * (1-uptake_program) + targetgroup * uptake_program * (1-vaccine_efficacy)

program_infections         <- population_at_risk * force_of_infection 
program_hospitalisations   <- program_infections * p_hosp
program_deaths             <- program_infections * case_fatality_ratio
program_lifeyearslost      <- program_deaths * life_expectancy
program_lifeyearslost_disc <- program_deaths * life_expectancy

program_treatment_cost     <- program_infections * treatment_cost_per_case
program_vaccination_cost   <- fixed_program_cost + targetgroup * uptake_program * vaccine_cost_per_dose
program_total_cost   <- program_treatment_cost + program_vaccination_cost

program_QALYloss          <- program_infections * QALYloss_per_case + program_lifeyearslost
program_QALYloss_disc     <- program_infections * QALYloss_per_case + program_lifeyearslost_disc

```

Calculate the burden of disease for the comparator without the program in place, including the medical costs:
```{r eval = FALSE}
### Similar to the previous tutorial: "HTA Flu (basic)"
```

```{r include = FALSE}
# Comparator
comparator_infections         <- targetgroup * force_of_infection
comparator_hospitalisations   <- comparator_infections * p_hosp
comparator_deaths             <- comparator_infections * case_fatality_ratio
comparator_lifeyearslost      <- comparator_deaths * life_expectancy
comparator_lifeyearslost_disc <- comparator_deaths * life_expectancy_disc
        
comparator_treatment_cost     <- comparator_infections * treatment_cost_per_case
comparator_vaccination_cost   <- 0
comparator_total_cost   <- comparator_treatment_cost + comparator_vaccination_cost

comparator_QALYloss       <- comparator_infections * QALYloss_per_case + comparator_lifeyearslost
comparator_QALYloss_disc  <- comparator_infections * QALYloss_per_case + comparator_lifeyearslost_disc

```

Calculate the difference between the program and the comparator:
```{r eval = FALSE}
### Similar to the previous tutorial: "HTA Flu (basic)"
```

```{r include = FALSE}
# To calculate the difference between 'program' and 'comparator' 
infections_averted          <- comparator_infections - program_infections
hospitalisations_averted    <- comparator_hospitalisations - program_hospitalisations
deaths_averted              <- comparator_deaths - program_deaths
lifeyearslost_averted       <- comparator_lifeyearslost - program_lifeyearslost
lifeyearslost_averted_disc  <- comparator_lifeyearslost_disc - program_lifeyearslost_disc

QALY_gained                 <- comparator_QALYloss - program_QALYloss
QALY_gained_disc            <- comparator_QALYloss_disc - program_QALYloss_disc

treatment_cost_averted     <- comparator_treatment_cost - program_treatment_cost
incr_vaccination_cost      <- program_vaccination_cost - comparator_vaccination_cost # incremental cost
incr_total_cost            <- incr_vaccination_cost - treatment_cost_averted

```

Calculate the incremental effectiveness ratios for infections, hospitalizations, and deaths averted. 
Warning, the expected ICER equals the expected incremental cost / average incremental effect. this is NOT the same as the average all ICER$_i$ values!! Finally, calculate the incremental effectiveness ratios for life years saved:

```{r }

# Calculate incremental effectiveness ratios
incr_cost_per_case_prevented            <- mean(incr_total_cost) / mean(infections_averted)
incr_cost_per_hospitalisation_prevented <- mean(incr_total_cost) / mean(hospitalisations_averted)
incr_cost_per_death_averted             <- mean(incr_total_cost) / mean(deaths_averted)
incr_cost_per_lifeyear_gained           <- mean(incr_total_cost) / mean(lifeyearslost_averted)
incr_cost_per_lifeyear_gained_disc      <- mean(incr_total_cost) / mean(lifeyearslost_averted_disc)
incr_cost_per_QALY_gained               <- mean(incr_total_cost) / mean(QALY_gained)
incr_cost_per_QALY_gained_disc          <- mean(incr_total_cost) / mean(QALY_gained_disc)

```

Show the results:
```{r }
# combine results
# note: specify column names at once
ICER_values<-cbind(incr_cost_per_case_prevented        = incr_cost_per_case_prevented,
               incr_cost_per_hospitalisation_prevented = incr_cost_per_hospitalisation_prevented,
               incr_cost_per_death_averted             = incr_cost_per_death_averted,
               incr_cost_per_lifeyear_gained           = incr_cost_per_lifeyear_gained,
               incr_cost_per_lifeyear_gained_disc      = incr_cost_per_lifeyear_gained_disc,
               incr_cost_per_QALY_gained               = incr_cost_per_QALY_gained,
               incr_cost_per_QALY_gained_disc          = incr_cost_per_QALY_gained_disc)

# print rounded results
round(ICER_values)

```
Please note that there may be slight variations due to platform and R-version specific differences in random number generation.

## Second thoughts on the distributions for costs and cases

Please note that the mean estimate for the incremental total cost per infection averted differs from the value we obtained when not accounting for uncertainty. This discrepancy could be due to extreme values in the ratio. Kindly review the following plots and estimates:

```{r echo=FALSE}
# Set up the plotting area to have 1 row and 3 columns
par(mfrow = c(1, 3))

# Plot the histograms
hist(incr_total_cost, xlab = 'Incremental total cost', main = 'Histogram of Incremental Total Cost')
hist(QALY_gained_disc, xlab = 'QALY gain', main = 'Histogram of Infections Averted')
hist(incr_total_cost / QALY_gained_disc, xlab = 'Incremental total cost per infection averted', main = 'Histogram of Incremental Cost per Infection Averted')

# Reset the plotting layout to default
par(mfrow = c(1, 1))
```

```{r}
mean(incr_total_cost/QALY_gained_disc)       # mean of ratios
median(incr_total_cost/QALY_gained_disc)     # median of ratios
mean(incr_total_cost)/mean(QALY_gained_disc) # ratio of means: ICER!!
```


# Cost-effectiveness analysis considering uncertainty (using BCEA package).

1. Run the `bcea()`function to obtain the cost-effectiveness plane, the expected incremental net monetary benefit (INMB), the cost-effectiveness acceptability curve (CEAC) and the expected value for perfect information (EVPI):
```{r }
# inspect the documentation of the 'bcea()' function
?bcea

# Specify a matrix containing the clinical effectiveness for each intervention being considered.
# note: this needs to be specified as health gain (i.e. negative health loss)
ce_effects <- cbind(-comparator_QALYloss_disc,-program_QALYloss_disc)


# Specify a matrix containing the cost for each intervention being considered.
# note: this needs to be specified as health gain (i.e. negative health loss)
ce_cost   <- cbind(comparator_total_cost,program_total_cost)

# Define the labels to be associated with each intervention.
interventions = c("No intervention","Vaccine program")

# Run the 'bcea()' function
m <- bcea(eff = ce_effects,
          cost = ce_cost,
          ref = 2,  # the column of eff and cost considered to be the reference strategy.
          interventions = interventions,
          plot = T) #ref specifies to compare 'Vaccine program' to 'No intervention'

```

2. Which decision should we take, given current information and for a range of WTP values? Print the summary of the `bcea()` function:
```{r }
summary(m)
```


3. How much evidence is there in favour of the decision, for a range of WTP values? Look at the CEAC and CEAF.
```{r }
m_ceaf <- multi.ce(m)
ceaf.plot(m_ceaf) # uses base plot 

```    

4. Is there value in further research? Look at the EVPI graph and summary above.
```{r }

```


<br>
<br>

<div>
# Navigate
Top: [Index](index.html)
</div>

