---
title: "HTA for RSV with static and dynamic disease model"
toc: FALSE
---

<div style="text-align: right">
[Home page](index.html)
</div>

```{r libs, cache=FALSE, echo=FALSE, results = "hide", warning = FALSE}
library('curl')
library('knitr')
library('EpiDynamics')
set.seed(2020)

knitr::opts_chunk$set(fig.cap='', fig.align="center", message=FALSE, warning=FALSE, fig.height = 4, fig.width = 6)
```

This tutorial has been developed by Joke Bilcke and Lander Willem, for illustration and teaching purposes only. 
Please make contact if you would like to use this model in any public way.

The burden of respiratory syncytial virus (RSV) in older adults resembles that of influenza and may even be considered worse due to the lack of preventive interventions (Zeevat 2022). In this tutorial we explore the cost-effectiveness of a hypothetical RSV vaccine for the adult population of in the Netherlands. We start from a static disease model, which is most applicable if vacccination targets a small proportion of the population that does not at the same time form a core transmitter group of the virus. In a later stage, we integrate a dynamic transmission model into the health economic evaluation.

For further information, questions or comments please contact: 
joke.bilcke@uantwerp.be or lander.willem@uantwerp.be from the 
Centre for Health Economics Research and Modeling of Infectious Diseases (CHERMID),
Vaccine and Infectious Disease Institute (Vaxinfectio),
University of Antwerp,
Universiteitsplein 1,
2610 Antwerp,
Belgium.

# Set working directory
The first step when creating a new R file is to specify a working directory. You need to tell the R session where you want it to work. To do this you use the command to "set the working directory" or `setwd()`. 

You can label this working directory as 'home', and have other locations stored for plotting or for getting data. This location should be an existing folder on your computer. For example

```{r results = "hide", eval = FALSE}
home = "~/Documents/Modelling_intro/" ## on a OS X
home = "C:\\Documents\\Modelling_intro\\" ## on windows
setwd(home)
```




# MODEL INPUT

Input parameters are inspired by the work of Zeevat et al (2022), published in the Journal of Infectious Diseases.

```{r }

# population and disease input
population_size           = 4557901 # NL population +60
rsv_incidence             = 0.03525 # in the absence of vaccination
rsv_recovery_time         = 7 # days

# health economic input
cost_per_episode          = 122 # euro
qaly_loss_per_episode     = 0.002196

# vaccine characteristics
vaccine_efficacy  = 0.40
cost_per_dose     = 10 # euro 
admin_cost        = 50 # euro

# program settings
uptake = 0.2 # doses
time_horizon = 365 # days

```

# PRE-PROCESSING
Based on the input above, we can compute the program cost and the proportion of the population that will be effectively protected:
```{r}
program_costs = population_size * uptake * (cost_per_dose + admin_cost)
proportion_protected =  uptake * vaccine_efficacy
```


# STATIC DISEASE MODEL
```{r }
# TRANSLATE TO STATIC MODEL SETTING
force_of_infection  = rsv_incidence

# INFECTIONS (prevented)
reference_infections = population_size * force_of_infection
program_infections = population_size * uptake * force_of_infection * (1-vaccine_efficacy) + 
                       population_size * (1 - uptake) * force_of_infection
prevented_cases = reference_infections - program_infections

# COSTS AND EFFECTS
static_incr_cost = program_costs - (prevented_cases * cost_per_episode)
static_qaly_gain = prevented_cases * qaly_loss_per_episode

# INCREMENTAL COST EFFECTIVENESS RATIO
static_icer = static_incr_cost  / static_qaly_gain

# PRINT RESULTS
round(c(reference_infections = reference_infections,
        program_infections = program_infections,
        static_qaly_gain = static_qaly_gain,
        static_icer_k = static_icer / 1e3),digits=2)

```




# DYNAMIC DISEASE MODEL: introduction

For the dynamic model, we make use of the [EpiDynamics](https://cran.r-project.org/web/packages/EpiDynamics/EpiDynamics.pdf) package with build in functions. For example, we can adopt the 'SIR' function:

```{r}
# check if the "EpiDynamics" package is installed, if not, install package
if(!'EpiDynamics' %in% installed.packages()){ install.packages('EpiDynamics') }

# load library
library(EpiDynamics)

# Parameters and initial conditions. 
parameters = list(beta = 5.147, gamma = 1 / rsv_recovery_time)
initials = c(S = 1 - 1e-06 - 0.2, I = 1e-06, R = 0.2) 

# Solve and plot. 
sir = SIR(pars = parameters, init = initials, time = 0:time_horizon) 

# Plot output
PlotMods(sir)
```

To define the number of infections, we can calculate the change in the "Recovered" compartment:
```{r}
sir$results$R[nrow(sir$results)] - sir$results$R[1]
```

# DYNAMIC DISEASE MODEL: RSV

To meet the target incidence of RSV for this CEA, we can assume a high existing immunity level in the population. 
```{r }
# set initial immunity levels
initials['R'] = 0.951
initials['S'] = 1 - sum(initials['I']+initials['R'])

# Solve 
dynamic_rsv_reference = SIR(pars = parameters,init = initials,time = 0:time_horizon)

# calculate total infections
dynamic_cases_reference = diff(dynamic_rsv_reference$results$R[c(1,nrow(dynamic_rsv_reference$results))]) * population_size
dynamic_cases_reference # print to terminal
```

To include vaccination in the model, we can consider vaccine uptake as transfer from "Suseptible" to "Recovered":

```{r }
initials['S'] = initials['S'] * (1-proportion_protected)
initials['R'] = 1 - sum(initials['S']+initials['I'])

# Solve 
dynamic_rsv = SIR(pars = parameters,init = initials,0:time_horizon)

#PlotMods(dynamic_rsv)
dynamic_cases_program = diff(dynamic_rsv$results$R[c(1,nrow(dynamic_rsv$results))]) * population_size

# prevented cases
dynamic_cases_prevented = dynamic_cases_reference - dynamic_cases_program

# CEA
dynamic_incr_cost = program_costs - (dynamic_cases_prevented * cost_per_episode)
dynamic_qaly_gain = dynamic_cases_prevented * qaly_loss_per_episode

# ICER
dynamic_icer = dynamic_incr_cost  / dynamic_qaly_gain

# PRINT RESULTS
round(c(dynamic_cases_reference=dynamic_cases_reference,
        dynamic_cases_program=dynamic_cases_program,
        dynamic_qaly_gain=dynamic_qaly_gain,
        dynamic_icer=dynamic_icer / 1e3),digits=2)

```

# DYNAMIC DISEASE MODEL: seasonality

Another option is to include seasonality and less (long-lasting) immunity. To do this, the 'SIRSinusoidalForcing' function can be used, in which the SIR model is extended with sinusoidal forcing of the transmission rate. 

We provide the parameters here:

```{r}

## EpiDynamics PACKAGE: seasonality
parameters = list(beta0 = 0.358385, 
                  beta1 = 3, 
                  gamma = 1 / rsv_recovery_time,
                  omega = 4 * pi / 365,
                  mu = 0) # no births/deaths at this stage

initials = c(S = 1 - 2e-5 - 0.8, 
             I = 2e-5,
             R = 0.8)

# Solve and plot.
seasonal_rsv_reference = SIRSinusoidalForcing(pars = parameters,
                                 init = initials,
                                 time = 0:time_horizon)

#PlotMods(seasonal_rsv_reference)
seasonal_cases_reference = diff(seasonal_rsv_reference$results$R[c(1,nrow(seasonal_rsv_reference$results))]) * population_size

```

To include vaccination in the model, we can consider vaccine uptake as transfer from "Suseptible" to "Recovered":
```{r}
initials['S'] = initials['S'] * (1-proportion_protected)
initials['R'] = 1 - sum(initials['S']+initials['I'])

# Solve
seasonal_rsv = SIRSinusoidalForcing(pars = parameters,
                                 init = initials,
                                 time = 0:(1 * 365))

#PlotMods(seasonal_rsv)
seasonal_cases_program = diff(seasonal_rsv$results$R[c(1,nrow(seasonal_rsv$results))]) * population_size

# prevented cases
seasonal_cases_prevented = seasonal_cases_reference - seasonal_cases_program

# CEA
seasonal_incr_cost = program_costs - (seasonal_cases_prevented * cost_per_episode)
seasonal_qaly_gain = seasonal_cases_prevented * qaly_loss_per_episode

# ICER
seasonal_icer =  seasonal_incr_cost / seasonal_qaly_gain

# print results
round(c(seasonal_cases_reference=seasonal_cases_reference,
        seasonal_cases_program=seasonal_cases_program,
        seasonal_qaly_gain=seasonal_qaly_gain,
        seasonal_icer_k=seasonal_icer / 1e3),digits=2)

```

# REPRODUCTON NUMBER
In an SIR model, the basic reproduction number is defined by the beta and gamma parameter. I.e., R_0 = beta * 1/gamma. We can use this for the seasonal implementation of 'beta' and calculate a theoretical reproduction number over time.
```{r}

# with the parameters of the "immunity model", the basic reproduction number is:
dynamic_rsv$pars$beta / dynamic_rsv$pars$gamma

# inspect seasonal RSV model:
seasonal_rsv$model
seasonal_rsv$pars

# i.e. the implementation of the seasonality:
# Please note that this formula assumes a fully susceptible population and 1 infected!!
# beta <- beta0 * (1 + beta1 * sin(omega * time))

# as such:
seasonal_beta = seasonal_rsv$pars$beta0 * (1 + seasonal_rsv$pars$beta1 * sin(seasonal_rsv$pars$omega * 0:time_horizon))

# plot "theoretical" basic reproduction number
plot(0:time_horizon,
     seasonal_beta * rsv_recovery_time,
     xlab='time (days)',
     ylab='reproduction number')
abline(h=1)

```


# SUMMARY

Inspect the cost-effectiveness frontier:
```{r}

plot(x = c(static_qaly_gain,dynamic_qaly_gain,seasonal_qaly_gain),
     y = c(static_incr_cost,dynamic_incr_cost,seasonal_incr_cost)/1e6,
     col = 1:3,
     xlab= 'QALY gain',
     ylab = 'incremental cost (euro)',
     pch=16)
legend('topright',
       c('static model',
         'dynamic model - immunity',
         'dynamic model - seasonality'),
       fill = 1:3,
       )
```

```{r, echo=F}

data.frame(model = c('static','dynamic - immunity', 'dynamic - seasonal'),
           reference_infections = c(reference_infections, 
                                    dynamic_cases_reference, 
                                    seasonal_cases_reference),
              program_infections = c(program_infections,
                                      dynamic_cases_program,
                                      seasonal_cases_program),
              qaly_gain = c(static_qaly_gain,dynamic_qaly_gain,seasonal_qaly_gain),
              program_icer_k = c(static_icer,dynamic_icer,seasonal_icer )/ 1e3) -> db_out

db_out[,-1] <- round(db_out[,-1])

db_out$infections_reduction <- round((db_out$reference_infections - db_out$program_infections) / reference_infections,digits=2)

kable(db_out)
 
```

Inspection of the disease prevalence:
```{r echo=F}


plot(x = dynamic_rsv_reference$time,
         y = dynamic_rsv_reference$results$I,
        xlab='time (days)',
        ylab='prevalence (infected)',
     ylim=c(0,max(c(dynamic_rsv_reference$results$I,seasonal_rsv_reference$results$I))),
        col=1,
        lwd=2,
        type='l')
lines(x = dynamic_rsv$time,
         y = dynamic_rsv$results$I,
        col=1,
        lwd=2,
      lty=2)

lines(x = seasonal_rsv_reference$time,
         y = seasonal_rsv_reference$results$I,
        col=4,
        lwd=2)
lines(x = seasonal_rsv$time,
         y = seasonal_rsv$results$I,
        col=4,
        lwd=2,
      lty=2)
legend('topright',
       c('dynamic - immunity: reference',
         'dynamic - immunity: program',
         'dynamic - seasonality: reference',
         'dynamic - seasonality: program'),
       lwd=2,
       col=c(1,1,4,4),
       lty=c(1,2,1,2),
       cex=0.8)
# lines(x = seasonal_rsv_reference$time,
#          y = rep(reference_infections/population_size/time_horizon,time_horizon+1),
#         col=5,
#         lwd=2)

```


# What if?

<n style="color:darkred">

* What happens to the ICERs if you change the vaccine uptake to 0.1 or 0.7? 
* Does it affect the outcome for all disease models? Why (not)?
* What happens to the ICERs if we combine RSV and Influenza vaccine uptake? (tip: "admission cost")
* Adapt the "immunity model" to start with an immunity level of 0.96387 and beta equals 15. What happens to the ICER, and why?
</n>

# Next...

<n style="color:darkred">

Let us assume the seasonal RSV incidence is 70%, while other model parameters remain unchanged. The following model parameters are aligned with the 70% incidence after one year. What are the corresponding ICERs? Can you explain the (new) differences? What about the reproduction numbers? 

</n>


```{r, eval=F}
# STATIC
force_of_infection = 0.7

# DYNAMIC: 70% INCIDENCE & NO EXISTING IMMUNITY
parameters = list(beta = 0.2458, gamma = 1 / rsv_recovery_time)
initials = c(S = 1 - 2e-5, I = 2e-5, R = 0)

# DYNAMIC: 70% INCIDENCE & EXISTING IMMUNITY
parameters = list(beta = 3.245, gamma = 1 / rsv_recovery_time)
initials = c(S = 1 - 2e-5 - 0.2999, I = 2e-5, R = 0.2999)

# DYNAMIC: 70% INCIDENCE & SEASONALITY 
parameters = list(beta0 = 0.11444, beta1 = 1.4, gamma = 1 / rsv_recovery_time, omega = 2 * pi / 365, mu = 0)
initials = c(S = 1 - 2e-5 , I = 2e-5, R = 0)

```


<div>
# Navigate
Top: [Index](index.html)
</div>

