---
title: "Demo Decision Tree (basic)"
toc: FALSE
---

(Page under construction)
```{r libs, cache=FALSE, echo=FALSE, results = "hide", warning = FALSE}
library('curl')
set.seed(2020)

knitr::opts_chunk$set(fig.cap='', fig.align="center", message=FALSE, warning=FALSE, fig.height = 6, fig.width = 10)
```

This model was developed by Philippe Beutels and Joke Bilcke, for illustration and teaching purposes only. 
Please make contact if you would like to use this model in any public way.

In the field of economic evaluation in general, it could be applicable to any intervention with effectiveness realised 
within one year (like many curative interventions). In the field of infectious disease, it would only be applicable to 
influenza, and only if vacccination targets a small proportion of the population that does not at the same time form a 
core transmitter group of the virus (eg it is unlikely to be suitable to model childhood influenza vaccination). Note 
that for most applications, practical use of this model would likely entail having separate, though similarly simple 
decision trees to obtain the unit cost estimates (implying that these cells would be intermediary 
outcomes, and that there would be additional input cells for each relevant stage of disease (for instance for the number
of consultations, the unit costs per consultation, medication use, hospital days etc for cases of pneumonia, and all 
other relevant disease stages)).

Note that the final output is independent of coverage and target group size, if fixed administration costs are set to 0. 
This is due to the exclusion of herd immunity effects in this model.  

For further information, questions or comments please contact: 
philippe.beutels@uantwerp.be or joke.bilcke@uantwerp.be
Centre for Health Economics Research and Modeling of Infectious Diseases (CHERMID)
Vaccine and Infectious Disease Institute (Vaxinfectio)
University of Antwerp
Universiteitsplein 1
2610 Antwerp
Belgium

# Set working directory
The first step when creating a new R file is to specify a working directory. You need to tell the R session where you want it to work. To do this you use the command to "set the working directory" or `setwd()`. 

You can label this working directory as 'home', and have other locations stored for plotting or for getting data. This location should be an existing folder on your computer. For example

```{r results = "hide", eval = FALSE}
home <- "~/Documents/Modelling_intro/" ## on a OS X
home <- "C:\\Documents\\Modelling_intro\\" ## on windows
setwd(home)
```


# INPUT
```{r }

targetgroup<-2500000 #people>65 years
uptake_program<-0.80
force_of_infection<-0.04
proportion_hospitalised<-0.10
casefatality_ratio<-0.03
life_expectancy<-12 #years
vaccine_efficacy<-0.60
vaccineprice_perdose<-10 #dollars
administrationcosts_perdose<-50 #dollars
cost_per_hospitalised_case<-2000 #dollars
cost_per_nonhospitalised_case<-100 #dollars
fixed_program_costs<-0 #dollars

```
# INTERMEDIARY OUTPUT
```{r }
variable_vaccinationcost_perdose<-vaccineprice_perdose+administrationcosts_perdose
treatment_costs_per_case<-cost_per_hospitalised_case*proportion_hospitalised+(1-proportion_hospitalised)*cost_per_nonhospitalised_case

#Program
program_infections<-targetgroup*uptake_program*force_of_infection*(1-vaccine_efficacy)+targetgroup*(1-uptake_program)*force_of_infection
program_hospitalisations<-program_infections*proportion_hospitalised
program_deaths<-program_infections*casefatality_ratio
program_lifeyearslost<-program_deaths*life_expectancy

program_treatmentcosts<-program_infections*treatment_costs_per_case
program_vaccinationcosts<-ifelse(uptake_program!=0,targetgroup*uptake_program*variable_vaccinationcost_perdose+fixed_program_costs,targetgroup*uptake_program*variable_vaccinationcost_perdose)
program_totaldirectcosts<-program_treatmentcosts+program_vaccinationcosts

#No Program
noprogram_infections<-targetgroup*force_of_infection
noprogram_hospitalisations<-noprogram_infections*proportion_hospitalised
noprogram_deaths<-noprogram_infections*casefatality_ratio
noprogram_lifeyearslost<-noprogram_deaths*life_expectancy

noprogram_treatmentcosts<-noprogram_infections*treatment_costs_per_case
noprogram_vaccinationcosts<-0
noprogram_totaldirectcosts<-noprogram_treatmentcosts+noprogram_vaccinationcosts

#Difference between 'program' and 'no program' (events and costs averted)
diff_infections<-noprogram_infections-program_infections
diff_hospitalisations<-noprogram_hospitalisations-program_hospitalisations
diff_deaths<-noprogram_deaths-program_deaths
diff_lifeyearslost<-noprogram_lifeyearslost-program_lifeyearslost

diff_treatmentcosts<-noprogram_treatmentcosts-program_treatmentcosts
diff_vaccinationcosts<-noprogram_vaccinationcosts-program_vaccinationcosts
diff_totaldirectcosts<--(diff_treatmentcosts+diff_vaccinationcosts)

```
# MAIN OUTPUT OF INTEREST
```{r }

incremental_costs_per_case_prevented<-diff_totaldirectcosts/diff_infections
incremental_costs_per_hospitalisation_prevented<-diff_totaldirectcosts/diff_hospitalisations
incremental_costs_per_death_averted<-diff_totaldirectcosts/diff_deaths
incremental_costs_per_lifeyear_gained<-diff_totaldirectcosts/diff_lifeyearslost

```

# OUTPUT TABLE
```{r }
ICER_names<-c('incrcosts per case prevented','incrcosts per hospitalisation prevented','incrcosts per death averted','incrcosts per life-year gained')
ICER_values<-c(incremental_costs_per_case_prevented,incremental_costs_per_hospitalisation_prevented,incremental_costs_per_death_averted,incremental_costs_per_lifeyear_gained)
names(ICER_values)<-ICER_names
round(ICER_values) #round for not showing decimals

```


<div>
# Navigate
Top: [Index](index.html)
</div>

