---
title: "Uncertainty distributions for the parameters of the flu model"
toc: FALSE
---

<div style="text-align: right">
[Home page](index.html)
</div>

```{r libs, cache=FALSE, echo=FALSE, results = "hide", warning = FALSE}
set.seed(2020)

knitr::opts_chunk$set(fig.cap='', fig.align="center", message=FALSE, warning=FALSE, fig.height = 6, fig.width = 10)

show_results = FALSE

```

This code is developed by Joke Bilcke and Lander Willem for illustration and teaching purposes only. 
For further information, questions or comments please contact us.

# Goal

Define the uncertainty distributions the following parameters of our flu model:

* Force of Infection: 4% (n=100)
* Proportion hospitalised: 10% (n=100)
* Case fatality ratio: 3% (n=1000)
* Life expectancy: 12 + 2 (mean + se)
* Vaccine efficacy: 60% [50%-70%]
* Administration costs per dose: 50 + 4 (mean + se)
* Cost per hospitalised case: $2000 + 100 (mean + se)
* Cost per non-hospitalised case: $100 + 5 (mean + se)
* QALY loss per hospitalised case: 0.018 + 0.0018 (mean + se)
* QALY loss per non-hospitalised case: 0.0082 + 0.0018 (mean + se)


# Set working directory
The first step when creating a new R file is to specify a working directory. You need to tell the R session where you want it to work. To do this you use the command to "set the working directory" or `setwd()`. 

You can label this working directory as 'home', and have other locations stored for plotting or for getting data. This location should be an existing folder on your computer. For example

```{r results = "hide", eval = FALSE}
home <- "~/Documents/Modelling_intro/" ## on a OS X
home <- "C:\\Documents\\Modelling_intro\\" ## on windows
setwd(home)
```

# Set number of samples
```{r }
nsamples = 10000

```

# FYI: random number seed
```{r }

set.seed(5698) # use this to get exactly the same results each time you run the code (e.g. for code checking)
sample(10)
sample(10)

# start over...
set.seed(5698) # use this to get exactly the same results each time you run the code (e.g. for code checking)
sample(10)
```
note: use `set.seed()` only once per script!


# Normal distributions
Let's first explore the parameter distribution if we assume the sample size large enough for the central theorem and there are no restrictions: sample from a normal distribution to include parameter uncertainty for the mean.

```{r }

set.seed(5698) # use this to get exactly the same results each time you run the code (e.g. for code checking)

# if probability (p) and sample size (n) given: se = sqrt(p*(1-p)/n
force_of_infection_norm      = rnorm(nsamples,0.04, sqrt(0.04 * (1-0.04) / 100))	
p_hospital_norm   = rnorm(nsamples,0.1, sqrt(0.1 * (1-0.1) / 100))		  
case_fatality_ratio_norm      = rnorm(nsamples,0.03, sqrt(0.03 * (1-0.03) / 1000)) 

# if mean and se given (and n unknown)
life_expectancy_norm = rnorm(nsamples,12,2)	

# if CI given: se = (CI_high - CI_low)/(2*1.96)
vaccine_efficacy_norm       = rnorm(nsamples,0.6, (0.7 - 0.5) / (2 * 1.96))			

# if mean and se given
adm_cost_norm       = rnorm(nsamples, 50, 4)		
cost_hosp_norm      = rnorm(nsamples, 2000, 100)
cost_non_hosp_norm  = rnorm(nsamples, 100, 5)	
QALYloss_hosp_case_norm      = rnorm(nsamples, 0.018, 0.0018) 
QALYloss_non_hosp_case_norm  = rnorm(nsamples, 0.0082, 0.0018)
```

# Other distributions
For some parameters, the sample size is rather small or they should be strictly positive, hence other distribution might be preferred. Let's first define some help functions to facilitate sampling using a given mean, se, proportion, samples size or confidence interval:

```{r}
sample_gamma <- function(nsamples, mean, se){
  rgamma(nsamples, shape = mean^2 / se^2, rate = mean / se^2)
}

sample_beta_event <- function(nsamples, events, sample_size){
  return(rbeta(nsamples, shape1 = events , shape2 = sample_size - events))
}

sample_beta_mean <- function(nsamples, mean, se){
  alpha = mean^2 * (1 - mean) / se^2 
  beta  = alpha * (1 - mean) / mean
  return(rbeta(nsamples, shape1 = alpha, shape2 = beta))
}
```


Let's sample from an adjusted distribution:

```{r , echo = !show_results, eval = FALSE}

# probability
force_of_infection_beta    = sample_beta_event(nsamples,4,100) # = rbeta(nsamples,4,96)

...

```

```{r , echo = show_results}

# probability
force_of_infection_beta    = sample_beta_event(nsamples,4,100)
p_hospital_beta            = sample_beta_event(nsamples,10,100)
case_fatality_ratio_beta   = sample_beta_event(nsamples,30,100)

# count data
life_expectancy_gamma =  sample_gamma(nsamples,12,2)

# efficacy =  probability 
vaccine_efficacy_beta = sample_beta_mean(nsamples, mean = 0.6, se = (0.7-0.5)/(2*1.96))

# efficacy = 1-RR
vaccine_efficacy_lognormal = 1 - exp(rnorm(nsamples,log(0.4),(log(0.5)-log(0.3))/(2*1.96)))
  
# cost = count data
adm_cost_gamma = sample_gamma(nsamples, 50, 4)
cost_hosp_gamma = sample_gamma(nsamples, 2000, 100)
cost_non_hosp_gamma = sample_gamma(nsamples, 100, 5)

# utility, not-negative: beta
QALYloss_hosp_case_beta = sample_beta_mean(nsamples, 0.018, 0.0018)

# utility, not-negative: beta
QALYloss_non_hosp_case_beta = sample_beta_mean(nsamples, 0.0082, 0.0018)

```

# Explore density plots

To check if sample size is sufficiently large to use normal distribution:

```{r, eval = TRUE}

plot(density(force_of_infection_norm), main = 'force of infection', ylim = c(0,25))
lines(density(force_of_infection_beta),col = 2)
abline(v = mean(force_of_infection_norm), lty = 2)
grid()
legend('topright',
      c('Norm()','Beta()', 'mean'),
      col = c(1,2,1),
      lty=c(1,1,2),
      lwd=1)
```

```{r, eval = show_results, echo = show_results}

plot(density(p_hospital_norm), main = 'probability of hospitalisation', ylim = c(0,25))
lines(density(p_hospital_beta),col = 2)
abline(v = mean(p_hospital_norm), lty = 2)
grid()
legend('topright',
      c('Norm()','Beta()', 'mean'),
      col = c(1,2,1),
      lty=c(1,1,2),
      lwd=1)


plot(density(case_fatality_ratio_norm), main = 'case fatality ratio')
lines(density(case_fatality_ratio_beta),col = 2)
abline(v = mean(case_fatality_ratio_norm), lty = 2)
grid()
legend('topright',
      c('Norm()','Beta()', 'mean'),
      col = c(1,2,1),
      lty=c(1,1,2),
      lwd=1)


plot(density(life_expectancy_norm), main = 'life expectancy')
lines(density(life_expectancy_gamma),col = 2)
abline(v = mean(life_expectancy_norm), lty = 2)
grid()
legend('topright',
      c('Norm()','Gamma()', 'mean'),
      col = c(1,2,1),
      lty=c(1,1,2),
      lwd=1)


plot(density(vaccine_efficacy_norm), main = 'vaccine effectiveness')
lines(density(vaccine_efficacy_lognormal),col = 2)
lines(density(vaccine_efficacy_beta),col = 3)
abline(v = mean(vaccine_efficacy_norm), lty = 2)
grid()
legend('topright',
      c('Norm()','logNorm()','Beta()', 'mean'),
      col = c(1,2,3,1),
      lty=c(1,1,1,2),
      lwd=1)

plot(density(adm_cost_norm), main = 'administration cost')
lines(density(adm_cost_gamma),col = 2)
abline(v = mean(adm_cost_norm), lty = 2)
grid()
legend('topright',
      c('Norm()','Gamma()', 'mean'),
      col = c(1,2,1),
      lty=c(1,1,2),
      lwd=1)

plot(density(cost_hosp_norm), main = 'cost per hospital case')
lines(density(cost_hosp_gamma),col = 2)
abline(v = mean(cost_hosp_norm), lty = 2)
grid()
legend('topright',
      c('Norm()','Gamma()', 'mean'),
      col = c(1,2,1),
      lty=c(1,1,2),
      lwd=1)

plot(density(cost_non_hosp_norm), main = 'cost per non-hospital case')
lines(density(cost_non_hosp_gamma),col = 2)
abline(v = mean(cost_non_hosp_norm), lty = 2)
grid()
legend('topright',
      c('Norm()','Gamma()', 'mean'),
      col = c(1,2,1),
      lty=c(1,1,2),
      lwd=1)

plot(density(QALYloss_hosp_case_norm), main = 'QALY loss per hospital case')
lines(density(QALYloss_hosp_case_beta),col = 2)
abline(v = mean(QALYloss_hosp_case_norm), lty = 2)
grid()
legend('topright',
      c('Norm()','Beta()', 'mean'),
      col = c(1,2,1),
      lty=c(1,1,2),
      lwd=1)

plot(density(QALYloss_non_hosp_case_norm), main = 'QALY loss per non-hospital case')
lines(density(QALYloss_non_hosp_case_beta),col = 2)
abline(v = mean(QALYloss_non_hosp_case_norm), lty = 2)
grid()
legend('topright',
      c('Norm()','Beta()', 'mean'),
      col = c(1,2,1),
      lty=c(1,1,2),
      lwd=1)

```

# Summary

```{r, echo = !show_results}
# over to you...
```

```{r, echo = show_results}

 
# * force of infection: sample size rather small for Normal distribution, use Beta
# * probability of hospitalisation: sample size rather small for Normal distribution, use Beta
# * case fatality ratio: sample size rather small for Normal distribution, use Beta
# 
# * life expectancy: Normal and Beta distribution OK
# * vaccine effectiveness: Normal, logNormal and Beta distributions OK		
# * administration cost: Normal and Gamma distributions OK
# * cost per hospital case: Normal and Gamma distributions OK
# * cost per non-hospital case: Normal and Gamma distribution OK
# * QALY loss per hospital case: Normal and Beta distribution OK
# 
# * QALY loss per non-hospital case: sample size rather small for normal distribution, use Beta
```



<div>
# Navigate
Top: [Index](index.html) ---
`r if (!show_results) {
    '<a href="HTA_flu_uncertainty_results.html"> Results</a>'
  } else {
    '<a href="HTA_flu_uncertainty.html"> Back</a>'
  }
`

</div>

