---
title: "CEA for RSV with static and dynamic disease models"
toc: FALSE
---

<div style="text-align: right">
[Home page](index.html)
</div>

```{r libs, cache=FALSE, echo=FALSE, results = "hide", warning = FALSE}
suppressPackageStartupMessages(library('curl'))
library('knitr')
library('EpiDynamics')
set.seed(2020)

knitr::opts_chunk$set(fig.cap='', fig.align="center", message=FALSE, warning=FALSE, fig.height = 4, fig.width = 6)
```

This tutorial has been developed by Joke Bilcke and Lander Willem, for illustration and teaching purposes only. 
Please make contact if you would like to use this model in any public way.
For further information, questions or comments please contact: 
joke.bilcke@uantwerp.be or lander.willem@uantwerp.be from the 
Centre for Health Economics Research and Modeling of Infectious Diseases (CHERMID),
Vaccine and Infectious Disease Institute (Vaxinfectio),
University of Antwerp,
Universiteitsplein 1,
2610 Antwerp,
Belgium.

# Research question 

In this session, we delve into a critical question: Is vaccinating the elderly against Respiratory Syncytial Virus (RSV) a cost-effective strategy when compared to not vaccinating them? RSV is known to cause lower respiratory tract infections in infants, the elderly, and individuals with underlying health conditions.

Amidst the backdrop of numerous emerging interventions aimed at preventing RSV, our focus today is to conduct a comprehensive evaluation of the cost-effectiveness of implementing a vaccination program for the elderly. To achieve this, we will employ both static and dynamic disease transmission models, allowing us to gain valuable insights into the potential impact of vaccination strategies.

While the ICER serves as a crucial measure of cost-effectiveness, we'll also encourage you to extend your analysis further. Consider calculating the Incremental Net Monetary Benefit (INMB) and the Net Loss (NL) associated with the vaccination program. This broader perspective will enhance your grasp of the economic implications and provide a more comprehensive view of the intervention's impact.

# Objectives of the Tutorial:

* Evaluate the cost-effectiveness of an elderly vaccination program against RSV in comparison to a scenario without such a program.
* Utilize disease transmission models to analyze the dynamics of RSV spread and its interactions within the population.
* Highlight the primary outcome measure of cost-effectiveness: the Incremental Cost-Effectiveness Ratio (ICER), quantified as the incremental cost per QALY gained.
* Calculate and interpret the INMB and/or the NL.

As we delve into the topic, we encourage you to explore various scenarios by manipulating input parameters. By doing so, you can gain a nuanced understanding of how altering these factors influences the ICER. This hands-on exploration will empower you to grasp the sensitivity of our analysis and its implications for decision-making.

# Set working directory
The first step when creating a new R file is to specify a working directory. You need to tell the R session where you want it to work. To do this you use the command to "set the working directory" or `setwd()`. 

You can label this working directory as 'home', and have other locations stored for plotting or for getting data. This location should be an existing folder on your computer. For example

```{r results = "hide", eval = FALSE}
home = "~/Documents/Modelling_intro/" ## on a OS X
home = "C:\\Documents\\Modelling_intro\\" ## on windows
setwd(home)
```



# MODEL INPUT

The disease incidence and economic parameters are loosely based on the work of Zeevat et al (2022), published in the Journal of Infectious Diseases.

```{r }

# population and disease input
population_size           = 4557901 # NL population +60
rsv_incidence             = 0.03525 # in the absence of vaccination
rsv_recovery_time         = 7 # days

# health economic input
cost_per_episode          = 122 # euro
qaly_loss_per_episode     = 0.002196

# vaccine characteristics
vaccine_efficacy  = 0.40
cost_per_dose     = 10 # euro 
admin_cost        = 50 # euro

# program settings
uptake = 0.2 # proportion of the population
time_horizon = 365 # days

```

# PRE-PROCESSING
Based on the input above, we can compute the program cost and the proportion of the population that will be effectively protected:
```{r}
program_costs = population_size * uptake * (cost_per_dose + admin_cost)
proportion_protected =  uptake * vaccine_efficacy
```


# STATIC DISEASE MODEL

We commence by delving into the static disease model. The following code delineates the model's intrinsic parameters and conducts ICER computation. The outcomes revolve around the baseline disease impact, portraying anticipated infections. Moreover, we explore infections within the intervention scenario, subsequently estimating the associated QALY improvement. This culmination leads us to the metric of cost-effectiveness: the ICER. This value is expressed in thousands of euros and rounded for clarity.

```{r }
# TRANSLATE TO STATIC MODEL SETTING
force_of_infection  = rsv_incidence

# INFECTIONS (prevented)
reference_infections = population_size * force_of_infection
program_infections = population_size * (1 - proportion_protected) * force_of_infection
prevented_cases = reference_infections - program_infections

# COSTS AND EFFECTS
static_incr_cost = program_costs - (prevented_cases * cost_per_episode)
static_qaly_gain = prevented_cases * qaly_loss_per_episode

# INCREMENTAL COST EFFECTIVENESS RATIO
static_icer = static_incr_cost  / static_qaly_gain

# PRINT RESULTS
round(c(reference_infections = reference_infections,
        program_infections = program_infections,
        static_qaly_gain = static_qaly_gain,
        static_icer = static_icer),digits=2)

```




# DYNAMIC DISEASE MODEL: general

Moving forward, our exploration advances to encompass the dynamic transmission model, while adhering to the identical model parameters and fundamental disease attributes as established in the static model.

For this example, we utilize the [EpiDynamics](https://cran.r-project.org/web/packages/EpiDynamics/EpiDynamics.pdf)  package, which features an array of built-in functions tailored for such analyses. A noteworthy example is the incorporation of the 'SIR' function, adeptly employed for quantifying the disease burden. Through this approach, we effectively calculate the disease's impact while considering distinct initial conditions and transmission parameters:


```{r}
# check if the "EpiDynamics" package is installed, if not, install package
if(!'EpiDynamics' %in% installed.packages()){ install.packages('EpiDynamics') }

# load library
library(EpiDynamics)

# Parameters and initial conditions. 
parameters = list(beta = 5.147, gamma = 1 / rsv_recovery_time)
initials = c(S = 1 - 1e-06 - 0.2, I = 1e-06, R = 0.2) 

# Solve and plot. 
sir = SIR(pars = parameters, init = initials, time = 0:time_horizon) 

# Plot output
PlotMods(sir)
```

To ascertain the quantity of infections, we can deduce this by evaluating the alteration in the "Recovered" compartment:
```{r}
sir$results$R[nrow(sir$results)] - sir$results$R[1]
```

# DYNAMIC DISEASE MODEL: RSV

In order to achieve the desired RSV incidence for this Cost-Effectiveness Analysis, we can posit a considerable existing level of immunity within the population:
```{r }
# set initial immunity levels
initials['R'] = 0.95099
initials['S'] = 1 - sum(initials['I']+initials['R'])

# Solve 
dynamic_rsv_reference = SIR(pars = parameters,init = initials,time = 0:time_horizon)

# calculate total infections
dynamic_cases_reference = diff(dynamic_rsv_reference$results$R[c(1,nrow(dynamic_rsv_reference$results))]) * population_size
dynamic_cases_reference # print to terminal
```

To incorporate vaccination within the model, we can conceptualize the adoption of the vaccine as a transition from the "Susceptible" to the "Recovered" compartment:

```{r }
initials['S'] = initials['S'] * (1-proportion_protected)
initials['R'] = 1 - sum(initials['S']+initials['I'])

# Solve 
dynamic_rsv = SIR(pars = parameters,init = initials,0:time_horizon)

#PlotMods(dynamic_rsv)
dynamic_cases_program = diff(dynamic_rsv$results$R[c(1,nrow(dynamic_rsv$results))]) * population_size

# prevented cases
dynamic_cases_prevented = dynamic_cases_reference - dynamic_cases_program

# CEA
dynamic_incr_cost = program_costs - (dynamic_cases_prevented * cost_per_episode)
dynamic_qaly_gain = dynamic_cases_prevented * qaly_loss_per_episode

# ICER
dynamic_icer = dynamic_incr_cost  / dynamic_qaly_gain

# PRINT RESULTS
round(c(dynamic_cases_reference=dynamic_cases_reference,
        dynamic_cases_program=dynamic_cases_program,
        dynamic_qaly_gain=dynamic_qaly_gain,
        dynamic_icer=dynamic_icer),digits=2)

```

Inspection of the disease prevalence:
```{r echo=F}


plot(x = dynamic_rsv_reference$time,
         y = dynamic_rsv_reference$results$I,
        xlab='time (days)',
        ylab='prevalence (infected)',
     ylim=c(0,max(dynamic_rsv_reference$results$I)),
        col=1,
        lwd=2,
        type='l')
lines(x = dynamic_rsv$time,
         y = dynamic_rsv$results$I,
        col=1,
        lwd=2,
      lty=2)

legend('topright',
       c('dynamic: reference',
         'dynamic: program'),
       lwd=2,
       col=c(1,1,4,4),
       lty=c(1,2,1,2),
       cex=0.8)

```

# SUMMARY

Inspect the cost-effectiveness frontier:
```{r}

plot(x = c(static_qaly_gain,dynamic_qaly_gain),
     y = c(static_incr_cost,dynamic_incr_cost),
     col = 1:2,
     xlab= 'QALY gain',
     ylab = 'incremental cost (euro)',
     pch=16)
legend('topright',
       c('static model',
         'dynamic model'),
       fill = 1:2,
       )
```

```{r, echo=F}

data.frame(model = c('static','dynamic'),
           reference_infections = c(reference_infections, 
                                    dynamic_cases_reference),
              program_infections = c(program_infections,
                                      dynamic_cases_program),
              qaly_gain = c(static_qaly_gain,dynamic_qaly_gain),
              program_icer = c(static_icer,dynamic_icer)) -> db_out

db_out[,-1] <- round(db_out[,-1])

db_out$infections_reduction <- round((db_out$reference_infections - db_out$program_infections) / reference_infections,digits=2)

kable(db_out)
 
```

# Incremental Net Monetary Benefit (INMB)

The INMB is a crucial measure used in health economic evaluations to assess the cost-effectiveness of different healthcare interventions. It quantifies the net economic gain achieved by adopting one intervention over another. Mathematically, the INMB is calculated as the difference between the incremental benefits and the incremental costs of two competing interventions. It takes into account both the effectiveness and the costs associated with each intervention. The INMB helps decision-makers understand the economic value of an intervention by weighing the additional benefits gained against the extra costs incurred when compared to an alternative.

Compute the Incremental Net Monetary Benefit for the RSV illustration provided, while adhering to a willingness-to-pay threshold of 20,000 euros per Quality-Adjusted Life Year (QALY).


# Exploring model behaviour and sensitivity

<n style="color:darkred">

1. **Impact of Vaccine Uptake:** Evaluate the effect of altering vaccine uptake to 0.1 or 0.7 on ICERs and INMB.

2. **Consistency Across Models:** Examine whether changes in vaccine uptake lead to similar outcomes in both disease models. Provide an explanation for the observed results.

3. **Combining Vaccines:** Investigate the consequences of combining RSV and Influenza vaccine uptake on ICERs. Consider the influence of admission costs in your analysis.

4. **Matching Static and Dynamic Models:** Identify the parameter settings within the dynamic model that would yield results consistent with the static model. Discuss the significance of these settings.

5. **The Role of Discounting:** Reflect on the decision not to incorporate discounting. Offer your recommendation on whether discounting should be added and provide a rationale.


</n>



<div>
# Navigate
Top: [Index](index.html)
</div>

