---
title: "Uncertainty distributions for the parameters of the flu model"
toc: FALSE
---

(Page under construction)
```{r libs, cache=FALSE, echo=FALSE, results = "hide", warning = FALSE}
library('curl')
set.seed(2020)

knitr::opts_chunk$set(fig.cap='', fig.align="center", message=FALSE, warning=FALSE, fig.height = 6, fig.width = 10)
```

This model was developed by Joke Bilcke, for illustration and teaching purposes only. 
For further information, questions or comments please contact: 
joke.bilcke@uantwerp.be
Centre for Health Economics Research and Modeling of Infectious Diseases (CHERMID)
Vaccine and Infectious Disease Institute (Vaxinfectio)
University of Antwerp
Universiteitsplein 1
2610 Antwerp
Belgium

# Set working directory
The first step when creating a new R file is to specify a working directory. You need to tell the R session where you want it to work. To do this you use the command to "set the working directory" or `setwd()`. 

You can label this working directory as 'home', and have other locations stored for plotting or for getting data. This location should be an existing folder on your computer. For example

```{r results = "hide", eval = FALSE}
home <- "~/Documents/Modelling_intro/" ## on a OS X
home <- "C:\\Documents\\Modelling_intro\\" ## on windows
setwd(home)
```

# Set number of samples
```{r }
nsamples=10000

```

# Normal uncertainty distributions
```{r }

set.seed(698) # use this to get exactly the same results each time you run the code (e.g. for code checking)
foi=rnorm(nsamples,4/100,sqrt(0.04*(1-0.04)/100))			#--> sample size too small for normal distr
prophosp=rnorm(nsamples,0.1,0.03)			#--> sample size too small for normal distr
casefat=rnorm(nsamples,0.03,0.003)  		#--> sample size too small for normal distr
lifexp=rnorm(nsamples,12,2)				#--> normal distr ok
ve=rnorm(nsamples,0.6,(0.7-0.5)/(2*1.96))		#--> normal distr ok		
admcost=rnorm(nsamples,50,4)				#--> normal distr ok
costhosp=rnorm(nsamples,2000,100)			#--> normal distr ok
costnonhosp=rnorm(nsamples,100,5)			#--> normal distr ok
qalyhosp=rnorm(nsamples,0.018,0.0018) 		#--> normal distr ok
qalynonhosp=rnorm(nsamples,0.0082,0.0018) 	#--> sample size slightly too low for normal distr

```

# Uncertainty distributions for small sample sizes
```{r }

foi2=rbeta(nsamples,4,96)
prophosp2=rbeta(nsamples,10,90)
casefat2=rbeta(nsamples,30,970)
lifexp2=rgamma(nsamples,12^2/2^2,12/2^2)
ve2_lognormal=1-exp(rnorm(nsamples,log(0.4),(log(0.5)-log(0.3))/(2*1.96)))
  ve2mean=0.6
  ve2se=(0.7-0.5)/(2*1.96)
  ve2alpha=((ve2mean**2*(1-ve2mean))/ve2se**2)-ve2mean
ve2_beta=rbeta(nsamples,ve2alpha,ve2alpha*(1-ve2mean)/ve2mean)
admcost2=rgamma(nsamples,50^2/4^2,50/4^2)
costhosp2=rgamma(nsamples,2000^2/100^2,2000/100^2)
costnonhosp2=rgamma(nsamples,100^2/5^2,100/5^2)
	qalyhosp2_alpha=(0.018^2*(1-0.018)/0.0018^2)-0.018
qalyhosp2=rbeta(nsamples,qalyhosp2_alpha,qalyhosp2_alpha*(1-0.018)/0.018)
	qalynonhosp2_alpha=(0.0082^2*(1-0.0082)/0.0018^2)-0.0082
qalynonhosp2=rbeta(nsamples,qalynonhosp2_alpha,qalynonhosp2_alpha*(1-0.0082)/0.0082)

```

# Density plots
```{r }
# To check if sample size is sufficiently large to use normal distribution
#par(mfrow=c(4,3))
plot(density(foi))
lines(density(foi2),col='red')
plot(density(prophosp))
lines(density(prophosp2),col='red')
plot(density(casefat))
lines(density(casefat2),col='red')
plot(density(lifexp))
lines(density(lifexp2),col='red')
plot(density(ve))
lines(density(ve2_lognormal),col='red')
lines(density(ve2_beta),col='green')
plot(density(admcost))
lines(density(admcost2),col='red')
plot(density(costhosp))
lines(density(costhosp2),col='red')
plot(density(costnonhosp))
lines(density(costnonhosp2),col='red')
plot(density(qalyhosp))
lines(density(qalyhosp2),col='red')
plot(density(qalynonhosp))
lines(density(qalynonhosp2),col='red')
par(mfrow=c(1,1))

```


<div>
# Navigate
Top: [Index](index.html)
</div>

