---
title: "Introduction to location-specific IBMs"
---

```{r libs, cache=FALSE, echo=FALSE, results = "hide", warning = FALSE}
library('curl')
set.seed(2020)
source('../lib/ibm_functions.R')

knitr::opts_chunk$set(fig.cap='', fig.align="center", message=FALSE, warning=FALSE, fig.height = 6, fig.width = 10)
```

# Objectives
The aim of this tutorial is to run scenario analyses with an individual-based model in R that accounts for location specific mixing.

In this session, you will

1. Run an IBM with household, school/work and community mixing.
2. Discuss the impact of transmission parameters and features

The best thing to do is to read each section and type (or copy and paste) the R commands (in grey boxes) into your own R session to check that you understand the code. <n style="color:darkred"> Hands-on questions are indicated in **red** </n>

# Modelling concepts

The individual-based model will make use of a virtual population in which each individual is part of a household, school/work (based on age), and the general community. Each day (= time step), individuals can interact at home, school/work and in the community. We specify for each location a contact probability or average number of contacts per day.

# Set working directory
The first step when creating a new R file is to specify a working directory. You need to tell the R session where you want it to work. To do this you use the command to "set the working directory" or `setwd()`. 

You can label this working directory as 'home', and have other locations stored for plotting or for getting data. For example

```{r results = "hide", eval = FALSE}
home <- "~/Documents/Modelling_intro/" ## on a OS X
home <- "C:\\Documents\\Modelling_intro\\" ## on windows
setwd(home)
```

# Load model code

Please load the IBM functions we prepared on the GitHub repository. Using your skills from the previous tutorial(s), you should be able to code this model yourselves.

```{r eval= FALSE}
# download R file with IBM function
curl_download("https://github.com/lwillem/modelling-intro/raw/master/lib/ibm_functions.R", destfile = "ibm_functions.R")
source('ibm_functions.R')
```

# Base case scenario

<n style="color:darkred">
First, lets run the model with the default parameters:
</n>

```{r }
# you can run the location-specific IBM with default parameters (see console)
run_ibm_location()

```

<n style="color:darkred">
What can you learn from these graphs?
</n>

# Change some parameters
```{r eval=FALSE}

# increase num_contacts_community_day to 6
run_ibm_location(num_contacts_community_day = 6)

# set num_schools to 5
run_ibm_location(num_school = 6)

# set transmission_prob to 0.15 (and switch off the demographic plots)
run_ibm_location(transmission_prob =  0.15,
                 bool_show_demographics = FALSE)

```


## Scenarios
<n style="color:darkred">
Now, what happens if you...

* increase the population size?
* start with more infected individuals?
* allow only household contacts?
* allow only community contacts?
* allow only school contacts?
* allow only household and school contacts?
* allow only household and workplace contacts?
* ... etc.
</n>

<div>
# Navigate
Top: [Index](index.html)
</div>

