---
title: "Location specific IBM"
---

```{r libs, cache=FALSE, echo=FALSE, results = "hide", warning = FALSE}
library('curl')
set.seed(2020)
source('../lib/ibm_functions.R')

knitr::opts_chunk$set(fig.cap='', fig.align="center", message=FALSE, warning=FALSE, fig.height = 6, fig.width = 10)
```

# Objectives
The aim of this tutorial is to run scenairo analyses with an individual-based model in R that accounts for location specific mixing.

In this session, you will

1. Run an IBM with household, school/work and community mixing.
2. Discuss the impact of transmission parameters and features

The best thing to do is to read each section and type (or copy and paste) the R commands (in grey boxes) into your own R session to check that you understand the code.

# Set working directory
The first step when creating a new R file is to specify a working directory. You need to tell the R session where you want it to work. To do this you use the command to "set the working directory" or `setwd()`. 

You can label this working directory as 'home', and have other locations stored for plotting or for getting data. For example

```{r results = "hide", eval = FALSE}
home <- "~/Documents/Modelling_intro/" ## on a OS X
home <- "C:\\Documents\\Modelling_intro\\" ## on windows
setwd(home)
```

# Load model code

You can use your model from the previous tutorial, or load an IBM function from the GitHub repository.

```{r eval= FALSE}
# download R file with IBM function
curl_download("https://github.com/lwillem/modelling-intro/raw/master/lib/ibm_functions.R", destfile = "ibm_functions.R")
source('ibm_functions.R')
```

# Scenario analyses

Now you have a working model, you can start changing input parameters and explore the effect on the transmission dynamics. First, lets run our model with the default parameters:

```{r }
# you can run the location-specific IBM with default parameters (see console)
run_ibm_location()

```

What can you learn from these graphs?


## Scenarios...
Now, what happens if you...

* increase the population size?
* decrease the area size?
* change the velocity?
* start with more infected indviduals?
* increase the contact range?
* increase the transmission probability?
* decrease the average number of contacts per day?
* change the time horizon?
* ... etc.


<div>
# Navigate
Top: [Index](index.html)
</div>

