---
title: "HTA RSV tutorials"
author: "lwillem"
date: "`r Sys.Date()`"
output: pdf_document
toc: TRUE
---

```{r setup, include=FALSE}
library('curl')
library('knitr')
library('EpiDynamics')
set.seed(2020)

knitr::opts_chunk$set(fig.cap='', fig.align="center", message=FALSE, 
                      warning=FALSE, fig.height = 7, fig.width = 7)

```

# Introduction
This document contains solutions of a tutorial on the cost-effectiveness of RSV vaccine strategies that has been developed by Joke Bilcke and Lander Willem, for illustration and teaching purposes only. Please make contact if you would like to use this model in any public way.

The primary aim of this exercise is to learn how to conduct a (simple) cost-utility analysis making use of a dynamic disease transmission model.
A secondary aim is to make use of generic functions to perform the cost-utility analysis for multiple scenarios. Using your skills from the previous tutorial(s), you should be able to code this model yourselves.

For further information, questions or comments please contact: 
joke.bilcke@uantwerp.be or lander.willem@uantwerp.be from the 
Centre for Health Economics Research and Modeling of Infectious Diseases (CHERMID),
Vaccine and Infectious Disease Institute (Vaxinfectio),
University of Antwerp,
Universiteitsplein 1,
2610 Antwerp,
Belgium.

\newpage
# Research question
Is vaccinating elderly against RSV cost-effective when compared to not vaccinating them? RSV refers to respiratory syncytial virus which causes lower respiratory tract infection in infants, elderly and persons with comorbidities. Currently, many new interventions are being tested for preventing RSV, the exercise for today is to evaluate if a vaccination program in elderly is cost-effective compared to no such program, using a static or a dynamic disease transmission model. The main outcome of cost-effective is the ICER expressed as incremental cost per QALY gained. In addition, try to explore different scenario’s (i.e. changing input parameters) and explore how this impacts on the ICER. You can also try to calculate the INMB and/or the NL.

# Default model input
The disease incidence and economic parameters are loosely based on the work of Zeevat et al (2022), published in the Journal of Infectious Diseases.

```{r }
# population and disease input
cea_input <- list(
  population_size           = 4557901, # NL population +60
  rsv_incidence             = 0.03525, # in the absence of vaccination
  rsv_recovery_time         = 7, # days
  
  # health economic input
  cost_per_episode          = 122, # euro
  qaly_loss_per_episode     = 0.002196,
  
  # vaccine characteristics
  vaccine_efficacy  = 0.40,
  cost_per_dose     = 10, # euro 
  admin_cost        = 50, # euro
  
  # program settings
  uptake       = 0.2, # doses
  time_horizon = 365 # days
)
```

\newpage
# Function library
## Static disease model
```{r}
run_cea_static = function(force_of_infection, cea_input){
  
  # INFECTIONS (prevented)
  cases_ref = cea_input$population_size * force_of_infection
  cases_vac = cea_input$population_size * cea_input$uptake * force_of_infection *
    (1 - cea_input$vaccine_efficacy) + cea_input$population_size * 
    (1 - cea_input$uptake) * force_of_infection
  
  # Get prevented cases
  prevented_cases = cases_ref - cases_vac
  
  # COSTS AND EFFECTS
  program_costs = cea_input$population_size * cea_input$uptake * 
    (cea_input$cost_per_dose + cea_input$admin_cost)
  incr_cost = program_costs - (prevented_cases * cea_input$cost_per_episode)
  qaly_gain = prevented_cases * cea_input$qaly_loss_per_episode
  
  # INCREMENTAL COST EFFECTIVENESS RATIO
  icer = incr_cost  / qaly_gain
  
  # PRINT RESULTS
  db_out = data.frame(incidence_ref = force_of_infection, cases_ref = cases_ref,
             cases_vac = cases_vac, qaly_gain = qaly_gain, incr_cost = incr_cost,
             icer_k = icer / 1e3)
  print(round(db_out,digits=2))
  
  return(db_out)
}
```

\newpage
## Dynamic transmission model

```{r eval= TRUE}
run_cea_dynamic = function(parameters, initials, cea_input) {
  
  # retrieve EpiDynamic function
  sir_function <- ifelse('beta0' %in% names(parameters),SIRSinusoidalForcing,SIR)
    
  # Solve (and plot)
  scen_ref = sir_function(pars = parameters,init = initials,time = 0:cea_input$time_horizon)
  #PlotMods(scen_ref)
  
  # Get relative and absolut incidence
  incidence_ref = diff(scen_ref$results$R[c(1,nrow(scen_ref$results))])
  cases_ref = incidence_ref * cea_input$population_size
  
  # include vaccination in the model by transfering individuals from "S" to "R":
  initials['S'] = initials['S'] * (1-(cea_input$uptake * cea_input$vaccine_efficacy))
  initials['R'] = 1 - sum(initials['S']+initials['I'])

  # Sove (and plot)
  scen_vac = sir_function(pars = parameters, init = initials, time = 0:cea_input$time_horizon)
  # PlotMods(scen_vac)
  
  # Get relative and absolut incidence
  program_incidence = diff(scen_vac$results$R[c(1,nrow(scen_vac$results))]) 
  cases_vac = program_incidence * cea_input$population_size
  
  # Get prevented cases, incremental cost and QALY gain
  prevented_cases = cases_ref - cases_vac
  program_costs = cea_input$population_size * cea_input$uptake * 
    (cea_input$cost_per_dose + cea_input$admin_cost)
  incr_cost = program_costs - (prevented_cases * cea_input$cost_per_episode)
  qaly_gain = prevented_cases * cea_input$qaly_loss_per_episode
  
  # calculate ICER
  icer = incr_cost / qaly_gain
  
  # Calculate R0
  if('beta0' %in% names(parameters)){
    seasonal_beta = parameters$beta0 * (1 + parameters$beta1 * 
                                          sin(parameters$omega * 0:cea_input$time_horizon))
    R0 =  seasonal_beta / parameters$gamma
  } else{ R0 = parameters$beta / parameters$gamma }

  # aggregate output
  db_out = data.frame(incidence_ref=incidence_ref, cases_ref=cases_ref, 
             cases_vac=cases_vac, qaly_gain=qaly_gain,incr_cost = incr_cost, 
             R0_min = min(R0), R0_max=max(R0), icer_k=icer / 1e3)
  print(round(db_out,digits=2))
  
  # return results
  return(db_out)
}
```

\clearpage
# CEA: scenario analysis

Explore CEA results from the online tutorial with default and adjusted parameters:

```{r}
# initial and adjusted cea_input parameter sets
cea_input_02 = cea_input_01 = cea_input_07 = cea_input_adm = cea_input
cea_input_01$uptake = 0.1 # 10% uptake
cea_input_07$uptake = 0.7 # 70% uptake
cea_input_adm$admin_cost = 0 # no (extra) admission costs
```

## Static model
```{r }
static_out_02  <- run_cea_static(force_of_infection = cea_input$rsv_incidence, cea_input_02)
static_out_01  <- run_cea_static(force_of_infection = cea_input$rsv_incidence, cea_input_01)
static_out_07  <- run_cea_static(force_of_infection = cea_input$rsv_incidence, cea_input_07)
static_out_adm <- run_cea_static(force_of_infection = cea_input$rsv_incidence, cea_input_adm)
```

## Dynamic model with high pre-existing immunity
```{r}
parameters = list(beta = 5.147, gamma = 1 / cea_input$rsv_recovery_time)
initials = c(S = 1 - 2e-5 - 0.951, I = 2e-5, R = 0.951)
immunity_out_02 <- run_cea_dynamic(parameters, initials, cea_input_02)
immunity_out_01 <- run_cea_dynamic(parameters, initials, cea_input_01)
immunity_out_07 <- run_cea_dynamic(parameters, initials, cea_input_07)
immunity_out_adm <- run_cea_dynamic(parameters, initials, cea_input_adm)
```

## Dynamic model with very high pre-existing immunity
```{r}
parameters = list(beta = 15, gamma = 1 / cea_input$rsv_recovery_time)
initials = c(S = 1 - 2e-5 - 0.96387, I = 2e-5, R = 0.96387)
immunity_out_02 <- run_cea_dynamic(parameters, initials, cea_input_02)
immunity_out_01 <- run_cea_dynamic(parameters, initials, cea_input_01)
immunity_out_07 <- run_cea_dynamic(parameters, initials, cea_input_07)
immunity_out_adm <- run_cea_dynamic(parameters, initials, cea_input_adm)
```

## Dynamic model with seasonality
```{r}
parameters = list(beta0 = 0.358385, beta1 = 3, gamma = 1 / cea_input$rsv_recovery_time, 
                  omega = 4 * pi / 365, mu = 0)
initials = c(S = 1 - 2e-5 - 0.8, I = 2e-5, R = 0.8)
seasonal_out_02 <- run_cea_dynamic(parameters, initials, cea_input_02)
seasonal_out_01 <- run_cea_dynamic(parameters, initials, cea_input_01)
seasonal_out_07 <- run_cea_dynamic(parameters, initials, cea_input_07)
seasonal_out_adm <- run_cea_dynamic(parameters, initials, cea_input_adm)
```

\newpage
## Graphical exploration:

```{r, eval=T}
plot(c(seasonal_out_01$qaly_gain,seasonal_out_02$qaly_gain,seasonal_out_07$qaly_gain),
     c(seasonal_out_01$incr_cost,seasonal_out_02$incr_cost,seasonal_out_07$incr_cost),
     xlab='QALY gain',ylab='incremental cost',xlim=c(0,350),ylim=c(0,2e8),pch=1:3,col=1:3,
     main=paste('SIR with seasonal forcing [incidence:',round(seasonal_out_01$incidence_ref,digits=3),']'))
lines(c(0,seasonal_out_01$qaly_gain),c(0,seasonal_out_01$incr_cost),col=1,lwd=2)
lines(c(0,seasonal_out_02$qaly_gain),c(0,seasonal_out_02$incr_cost),col=2,lwd=2)
lines(c(0,seasonal_out_07$qaly_gain),c(0,seasonal_out_07$incr_cost),col=3,lwd=2)
legend('topleft',
       c('uptake: 30%', 'uptake: 20%', 'uptake: 10%'),
       pch=3:1,col=3:1)
```


\clearpage
# CEA: counter-factual (force of infection = 70%)

## Static model
```{r }
static_out <- run_cea_static(force_of_infection = 0.7, cea_input)
```

## Dynamic model with NO pre-existing immunity
```{r }
parameters = list(beta = 0.2458, gamma = 1 / cea_input$rsv_recovery_time)
initials = c(S = 1 - 2e-5, I = 2e-5, R = 0)
dynamic_out <- run_cea_dynamic(parameters, initials, cea_input)
```

## Dynamic model with pre-existing immunity
```{r}
parameters = list(beta = 3.245, gamma = 1 / cea_input$rsv_recovery_time)
initials = c(S = 1 - 2e-5 - 0.2999, I = 2e-5, R = 0.2999)
immunity_out <- run_cea_dynamic(parameters, initials, cea_input)
```

## Dynamic model with seasonality
```{r}
parameters = list(beta0 = 0.11444, beta1 = 1.4, gamma = 1 / cea_input$rsv_recovery_time, 
                  omega = 2 * pi / 365, mu = 0)
initials = c(S = 1 - 2e-5 , I = 2e-5, R = 0)
seasonal_out <- run_cea_dynamic(parameters, initials, cea_input)
```

\newpage
```{r}
plot(x = c(static_out$qaly_gain,dynamic_out$qaly_gain,immunity_out$qaly_gain,seasonal_out$qaly_gain),
     y = c(static_out$incr_cost,dynamic_out$incr_cost,immunity_out$incr_cost,seasonal_out$incr_cost),
     pch=1:4, col=1:4, lwd=3,
     xlab= 'QALY gain',
     ylab = 'incremental cost (euro)')
abline(h=0,lty=3)
legend('topright',
       c('static model',
         'dynamic model - no previous immunity',
         'dynamic model - existing immunity',
         'dynamic model - seasonality'),
       col = 1:4, pch=1:4, lwd=0, cex=0.8
       )

```

\newpage
# Take home messages

* Static disease models do not take herd immunity effects into account. Only personal protection is considered in the economic evaluation. This holds when the target group has almost no effect on the transmission dynamics. The ICER does not change when the vaccine uptake varies. 

* Adding a new vaccine to an existing vaccination program, hence reducing the cost of the program, is beneficial for the cost-effectiveness of the vaccination program compared to no program.

* Dynamic disease models can enable herd immunity effects. As such, the number of prevented cases can outnumber the (effectively) protected individuals. This can result in a lower ICER as compared to when using a static model, since the health gains accounted for are higher for the same investment. Note that dynamic model can also enable other effects like an age shift (average age of infection increases). If infection at older age results in more severe illness, this can increase the ICER as compared to when using a static model, since the health gains can become lower. E.g. rubella vaccination of infants without catch-up at moderate uptake in Greece.  

* If the (basic) reproduction number is very high, vaccine-related immunization has limited impact on the effective transmission rate, hence herd immunity effects are limited. As such, the CEA results are more in line with the CEA based on the static disease model.

* If the (basic) reproduction number is close(r) to one, vaccine-related immunization can tip the balance and result in large herd immunity effects, hence resulting in CEA results more in favor of the program. Especially with the seasonal model, the high reproduction number is limited in time and reduced transmission seem to have most effect.

* With the dynamic model used for this example, we observed an inverse relationship between the ICER and vaccine uptake: the more vaccines are administered (i.e. the higher the uptake), the lower the ICER becomes. This is caused by the herd immunity effect. With low uptake resulting in high herd immunity, illness can be prevented 'for free' for the cases prevented due to herd immunity. 





