---
title: "Location-specific IBM tutorials"
author: "lwillem"
date: "`r Sys.Date()`"
output: pdf_document
---


```{r setup, include=FALSE}
library('curl')

set.seed(2020)
source('lib/ibm_functions.R')

knitr::opts_chunk$set(fig.cap='', fig.align="center", message=FALSE, 
                      warning=FALSE, fig.height = 6, fig.width = 10)

# set default: echo = false
#knitr::opts_chunk$set(echo = FALSE)


```
# Load model code

Please load the IBM functions we prepared on the GitHub repository. Using your skills from the previous tutorial(s), you should be able to code this model yourselves.

If the `curl` packages is not installed yet, you can do so by:
```{r results = "hide", eval=FALSE}
install.packages(curl) # install the curl library (to import data)
```

```{r eval= FALSE}
# load package 'curl'
library('curl')

# download R file with IBM function
curl_download("https://github.com/lwillem/modelling-intro/raw/master/lib/ibm_functions.R", destfile = "ibm_functions.R")

# load the IBM functions
source('ibm_functions.R')
```

\clearpage
# Base case scenario

```{r }
# you can run the location-specific IBM with default parameters (see console)
run_ibm_location()

```

\clearpage
# Population size 100
```{r pop_size_small, eval=TRUE}
run_ibm_location(pop_size = 100,
                 bool_show_demographics = FALSE,
                 add_baseline_prevalence = T)

```

\clearpage
# Population size 2000
```{r pop_size_large, eval=TRUE}
run_ibm_location(pop_size = 2000,
                 bool_show_demographics = FALSE,
                 add_baseline_prevalence = T)

```

\clearpage
# Start with more infected individuals
```{r num_infected_seeds, eval=TRUE}
run_ibm_location(num_infected_seeds = 100,
                 bool_show_demographics = FALSE,
                 add_baseline_prevalence = T)

```

\clearpage
# Add schools
```{r num_schools, eval=TRUE}
run_ibm_location(num_schools = 10,
                 bool_show_demographics = TRUE,
                 add_baseline_prevalence = T)

```

\clearpage
# Decrease the number of workplaces
```{r num_workplaces, eval=TRUE}
run_ibm_location(num_workplaces = 20,
                 bool_show_demographics = TRUE,
                 add_baseline_prevalence = T)

```

\clearpage
# Allow only household contacts
```{r only_hh, eval=TRUE}
run_ibm_location(num_contacts_community_day = 0,
                 contact_prob_household = 1,
                 contact_prob_school = 0,
                 contact_prob_workplace = 0,
                 bool_show_demographics = FALSE, add_baseline_prevalence = T)

```

\clearpage
# Allow only household and community contacts
```{r hh_community, eval=TRUE}
run_ibm_location(num_contacts_community_day = 5,
                 contact_prob_household = 1,
                 contact_prob_school = 0,
                 contact_prob_workplace = 0,
                 bool_show_demographics = FALSE, add_baseline_prevalence = T)

```

\clearpage
# Allow only school contacts
```{r only_school, eval=TRUE}
run_ibm_location(num_contacts_community_day = 0,
                 contact_prob_household = 0,
                 contact_prob_school = 1,
                 contact_prob_workplace = 0,
                 bool_show_demographics = FALSE, add_baseline_prevalence = T)

```

\clearpage
# Allow only household and school contacts
```{r hh_school, eval=TRUE}
run_ibm_location(num_contacts_community_day = 5,
                 contact_prob_household = 1,
                 contact_prob_school = 0.5,
                 contact_prob_workplace = 0,
                 bool_show_demographics = FALSE, add_baseline_prevalence = T)

```

\clearpage
# Allow only household and workplace contacts
```{r hh_workplace, eval=TRUE}
run_ibm_location(num_contacts_community_day = 0,
                 contact_prob_household = 1,
                 contact_prob_school = 0,
                 contact_prob_workplace = 0.5,
                 bool_show_demographics = FALSE, add_baseline_prevalence = T)

```

