---
title: "Random walk IBM tutorials"
author: "lwillem"
date: "`r Sys.Date()`"
output: pdf_document
---


```{r setup, include=FALSE}
library('curl')

set.seed(2020)
source('lib/ibm_functions.R')

knitr::opts_chunk$set(fig.cap='', fig.align="center", message=FALSE, 
                      warning=FALSE, fig.height = 6, fig.width = 10)

# set default: echo = false
#knitr::opts_chunk$set(echo = FALSE)


```
# Load model code

Please load the IBM functions we prepared on the GitHub repository. Using your skills from the previous tutorial(s), you should be able to code this model yourselves.

If the `curl` packages is not installed yet, you can do so by:
```{r results = "hide", eval=FALSE}
install.packages(curl) # install the curl library (to import data)
```

```{r eval= FALSE}
# load package 'curl'
library('curl')

# download R file with IBM function
curl_download("https://github.com/lwillem/modelling-intro/raw/master/lib/ibm_functions.R", destfile = "ibm_functions.R")

# load the IBM functions
source('ibm_functions.R')
```

\clearpage
# Base case scenario

```{r }
# you can run the location-specific IBM with default parameters (see console)
run_ibm_random_walk(rng_seed = 2000)

```

\clearpage
# Start with 10 infected individuals (instead of 3)
```{r eval=TRUE}
run_ibm_random_walk(num_infected_seeds = 10,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```

\clearpage
# Start with 50 infected individuals (instead of 3)
```{r eval=TRUE}
run_ibm_random_walk(num_infected_seeds = 50,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```


\clearpage
# number of days infected 3 (instead of 7)
```{r eval=TRUE}
run_ibm_random_walk(num_days_infected = 3,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```


\clearpage
# number of days infected 10 (instead of 7)
```{r eval=TRUE}
run_ibm_random_walk(num_days_infected = 10,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```

\clearpage
# Population size 100 (instead of 1000)
```{r pop_size_small, eval=TRUE}
run_ibm_random_walk(pop_size = 100,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```

\clearpage
# Population size 2000 (instead of 1000)
```{r pop_size_large, eval=TRUE}
run_ibm_random_walk(pop_size = 2000,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```

\clearpage
# Area size 10 (instead of 20)
```{r eval=TRUE}
run_ibm_random_walk(area_size = 10,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```

\clearpage
# Area size 40 (instead of 10)
```{r eval=TRUE}
run_ibm_random_walk(area_size = 40,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```

\clearpage
# Max contact distance 1 (instead of 2)
```{r eval=TRUE}
run_ibm_random_walk(max_contact_distance = 1,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```

\clearpage
# Max contact distance 4 (instead of 2)
```{r eval=TRUE}
run_ibm_random_walk(max_contact_distance = 4,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```

\clearpage
# Max velocity 0 (instead of 1)
```{r eval=TRUE}
run_ibm_random_walk(max_velocity = 0,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```

\clearpage
# Max velocity 4 (instead of 1)
```{r eval=TRUE}
run_ibm_random_walk(max_velocity = 4,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```

\clearpage
# transmission probability 0.05 (instead of 0.1)
```{r eval=TRUE}
run_ibm_random_walk(transmission_prob = 0.05,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```

\clearpage
# transmission probability 0.4 (instead of 0.1)
```{r eval=TRUE}
run_ibm_random_walk(transmission_prob = 0.4,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```

\clearpage
# vaccine coverage 50% (uniform)
```{r eval=TRUE}
run_ibm_random_walk(vaccine_coverage = 0.5,
                    apply_spatial_vaccine_refusal = F,
                    max_velocity = 0,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```

\clearpage
# vaccine coverage 50% (clustered)
```{r eval=TRUE}
run_ibm_random_walk(vaccine_coverage = 0.5,
                    apply_spatial_vaccine_refusal = T,
                    max_velocity = 0,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```

\clearpage
# vaccine coverage 80% (uniform)
```{r eval=TRUE}
run_ibm_random_walk(vaccine_coverage = 0.8,
                    apply_spatial_vaccine_refusal = F,
                    max_velocity = 0,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```

\clearpage
# vaccine coverage 80% (clustered)
```{r eval=TRUE}
run_ibm_random_walk(vaccine_coverage = 0.8,
                    apply_spatial_vaccine_refusal = T,
                    max_velocity = 0,
                    rng_seed = 2000,
                    add_baseline_prevalence = T)

```
